Initial spatial conditions in simulation models: the missing leg of sensitivity analyses?

### 2017-07-20 paper.tex
.

An urban system consists of a large number of social agents that interact with each other at a microscopic level in a system-time that evolves irreversibly creating temporal and cumulative effects. Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,Portugali2000}, and emergent properties at qualitatively differentiated scales~\citep{AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002,Batty2007}. In that frame, complexity is partially due to bifurcations \citep{Wilson1981} which are determinant in spatial systems~\citep{Wilson2002}. In spatially explicit simulation models, the non-linearity of local interactions is therefore very likely to sublimate small perturbations in the initial spatial setting, making it difficult to interprete the resulting global structures. \comment[JR]{je suis pour garder le passage sur la non-ergodicité (commenté ci-après) ; ça vous parait trop abstrait ?}\comment[MLT]{Non, mais je pense que la seconde phrase pourrait être reformulée/développée pour mieux raccrocher avec le reste, qu'en penses-tu?}
 % A crucial aspect of most spatio-temporal complex systems is their non-ergodicity~(\cite{pumain2012urban}) (the property that cross-sectional samples in space are not equivalent to samples in time to compute statistics such as averages), what witnesses generally strong spatio-temporal path-dependencies in their trajectories. Similar to what Gell-Mann calls \emph{frozen accidents} in any complex system~\cite{gell1995quark}, a given configuration contains clues on past bifurcations, that can have had dramatic effects on the state of the system.

>>>>

An urban system consists of a large number of social agents that interact with each other at a microscopic level in a system-time that evolves irreversibly creating temporal and cumulative effects. Self-organization has been shown to be a central feature of urban systems~\citep{AllenSanglier1981,Portugali2000}, and emergent properties at qualitatively differentiated scales~\citep{AzizAlaouiBertelle2009} must be obtained through simulation methods~\citep{Wu2002,Batty2007}. In that frame, complexity is partially due to bifurcations \citep{Wilson1981} which are determinant in spatial systems~\citep{Wilson2002}. In spatially explicit simulation models, the non-linearity of local interactions is therefore very likely to sublimate small perturbations in the initial spatial setting, making it difficult to interpret the resulting global structures. \comment[JR]{je suis pour garder le passage sur la non-ergodicité (commenté ci-après) ; ça vous parait trop abstrait ?}\comment[MLT]{Non, mais je pense que la seconde phrase pourrait être reformulée/développée pour mieux raccrocher avec le reste, qu'en penses-tu?}
 % A crucial aspect of most spatio-temporal complex systems is their non-ergodicity~(\cite{pumain2012urban}) (the property that cross-sectional samples in space are not equivalent to samples in time to compute statistics such as averages), what witnesses generally strong spatio-temporal path-dependencies in their trajectories. Similar to what Gell-Mann calls \emph{frozen accidents} in any complex system~\cite{gell1995quark}, a given configuration contains clues on past bifurcations, that can have had dramatic effects on the state of the system.
